---
title: "721: Project 2"
author: "Caroline Falvey"
date: '2022-10-04'
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

#Load libraries
library(readr)
library(dplyr)
library(lubridate)
library(here)
library(knitr)
library(gtsummary)
library(here)

#Clear lists
rm(list = ls())

```

## 

```{r, echo = TRUE, message = FALSE}

here::here("Duke", "Fall 2022", "721", "Project 2")

# Read in the two datasets
patient <- read_csv("Patient Level Data.csv")
encounter <- read_csv("Encounter Level Data.csv")

# Join the two datasets by MRN to create the full dataset
full_set <- full_join(patient, encounter, by = c("MRN"))

```


# Data Set Description:

The dataset contains `r {length(variable.names(full_set))}` variables with `r {length(full_set$contact_date)}` total encounters across `r {length(unique(full_set$MRN))}` patients. A full list of variables is included below:\

* *`r {variable.names(full_set)[1]}`*: a nominal variable containing each patient's medical record number (MRN) as a unique ID\
* *`r {variable.names(full_set)[2]}`*: a date variable containing each patient's date of birth\
* *`r {variable.names(full_set)[3]}`*: a categorical variable indicating each patient's race as white, black, or other\
* *`r {variable.names(full_set)[4]}`*: a categorical variable indicating whether each patient has Medicare or private insurance\
* *`r {variable.names(full_set)[5]}`*: a binary categorical variable classifying each patient's ethnicity as Hispanic or non-Hispanic\
* *`r {variable.names(full_set)[6]}`*: a binary categorical variable indicating if a patient was hypertensive during a given encounter\
* *`r {variable.names(full_set)[7]}`*: a binary categorical variable indicating if a patient has Congestive Heart Failure\
* *`r {variable.names(full_set)[8]}`*: a binary categorical variable indicating if a patient has diabetes\
* *`r {variable.names(full_set)[9]}`*: a date variable indicating the date of each encounter between 2014 and 2019\
* *`r {variable.names(full_set)[10]}`*: a binary categorical variable indicating if each encounter was a hospital visit or an office visit\
* *`r {variable.names(full_set)[11]}`*: a continuous variable with the patient's temperature (°F) for each encounter. Values range from `r {round(min(full_set$temp, na.rm = TRUE), 2)}`°F to `r {round(max(full_set$temp, na.rm = TRUE), 2)}`°F\
* *`r {variable.names(full_set)[12]}`*: an ordinal variable indicating a patient's distress score for each encounter. Values range from `r {min(full_set$distress_score, na.rm = TRUE)}` to `r {max(full_set$distress_score, na.rm = TRUE)}`\
* *`r {variable.names(full_set)[13]}`*: a continuous variable indicating each patient's white blood cell count (k/uL) at each encounter. Values range from `r {round(min(full_set$WBC, na.rm = TRUE), 2)}` k/uL to `r {round(max(full_set$WBC, na.rm = TRUE), 2)}` k/uL\
* *`r {variable.names(full_set)[14]}`*: a continuous variable indicating each patient's body mass index (BMI) at each encounter\

<br>

# Data Cleaning:

```{r data cleaning, echo = TRUE, results = "hide"}

# Create new variable for clean BMI data
full_set$BMI.r_clean <- full_set$BMI.r
# Truncate BMIs > 100 & <10
full_set$BMI.r_clean[full_set$BMI.r_clean > 100] <- 100
full_set$BMI.r_clean[full_set$BMI.r_clean < 10] <- 10

# Remove NAs in BMI data
full_set$BMI.r_clean[full_set$BMI.r_clean == "-999"] <- NA

# Create new variable for clean WBC data
full_set$WBC_clean <- full_set$WBC
# Truncate WBC < 0.05 k/uL
full_set$WBC_clean[full_set$WBC_clean < 0.05] <- 0.05

# Create new DOB variable using lubridate
full_set$DOB_clean <- mdy(full_set$DOB)
# Check for nonsensical DOBs
table(year(full_set$DOB_clean))
# Remove DOBs before 1900
full_set$DOB_clean[year(full_set$DOB_clean) < 1900] <- NA

# Remove raw data for DOB, WBC, and BMI from full dataset to use only cleaned variables
full_set <- full_set %>%
  select(-DOB, -WBC, -BMI.r)

```

<br>

# Temperature: 

```{r temperature, echo = TRUE}
# Round temp to one decimal place
full_set$rounded_temp <- round(full_set$temp, 1)

# Display first 10 values
full_set$rounded_temp[1:10]

```
<br> 

# WBC Categorization:

```{r WBC, echo = TRUE}

# Set ranges for WBC counts from Low - High in new variable
full_set$WBC_cat <- ifelse(full_set$WBC_clean < 3.2, "Low",
  ifelse(full_set$WBC_clean <= 9.8, "Normal",
    ifelse(full_set$WBC_clean > 9.8, "High",
      NA
    )
  )
)

# Set WBC_cat equal to Not Taken for NA values
full_set$WBC_cat[is.na(full_set$WBC_clean)] <- "Not Taken"

# Display results in logical order in a table
kable(table(full_set$WBC_cat)[c(1, 3, 2, 4)],
  align = "c",
  col.names = c("WBC Range", "Frequency")
)

```
<br>

# Elevated Temperature:

```{r elevated temp flag, ECHO = TRUE}

elv_temp <- full_set %>%
  select(MRN, contact_date, rounded_temp) %>% # select only MRN, contact date, and temp variables from full dataset
  filter(rounded_temp > 100) %>% # remove data for temp < 100
  arrange(MRN) # organize by MRN


# Display table of encounters with elevated temp (organized by MRN)
kable(elv_temp, align = "c", col.names = c("MRN", "Date", "Temperature"))
```

<br>

# Mean BMIs:

```{r BMI table, ECHO = TRUE}

bmi_table <- full_set %>%
  select(MRN, BMI.r_clean) %>% # select only MRN & BMI from the full dataset
  filter(MRN == "CI6950" |
    MRN == "IW9164" |
    MRN == "HJ8458" |
    MRN == "XE4615") %>% # filter by chosen MRNs
  group_by(MRN) %>%
  summarise(BMI = mean(BMI.r_clean, na.rm = TRUE)) # calculate mean for each MRN

# Round BMI to one decimal place
bmi_table$BMI <- round(bmi_table$BMI, 1)

# Display table of mean BMI by MRN
kable(bmi_table, align = "c")

```

<br>

# Number of Hospital Encounters per Year:

```{r hopital encounters, ECHO = TRUE}

# Convert format of contact date to year-month-date
full_set$contact_date_clean <- mdy(full_set$contact_date)

# Create new data set for just encounter types & contact dates
year_set <- full_set %>%
  select(enc_type, contact_date_clean) %>%
  filter(enc_type == "Hospital Encounter") # filter to just have hospital encounters

# Create new variable to store years of hospital encounters
cont_years <- year(year_set$contact_date_clean)

# Create a table showing how many hospital encounters occurred each year
kable(table(cont_years),
  align = "c",
  col.names = c("Year", "Hospital Encounters")
)

```

<br>

# Patient Ages:

```{r age, echo = TRUE}
# Create start date of 9/1/2019:
start_date <- as.Date("9/1/2019", tryFormats = "%m/%d/%Y")

# Calculate age using start date and DOB_clean variable created in Chunk 4: data cleaning
full_set$age <- as.numeric(difftime(
  time1 = start_date,
  time2 = full_set$DOB_clean,
  units = "days"
) / 365)

# Remove duplicated patients by filtering by duplicated MRN
age_set <- full_set %>%
  filter(!duplicated(MRN)) %>%
  select(MRN, age)

#Compute and display summary statistics for age:
age_set %>%
  summarise(
    Min = min(age, na.rm = TRUE),
    Q1 = quantile(age,
      probs = 0.25,
      names = FALSE, na.rm = TRUE
    ),
    Median = median(age, na.rm = TRUE),
    Mean = mean(age, na.rm = TRUE),
    Q3 = quantile(age,
      probs = 0.75,
      names = FALSE, na.rm = TRUE
    ),
    Max = max(age, na.rm = TRUE),
    SD = sd(age_set$age, na.rm = TRUE)
  ) %>%
  kable(digits = 2, align = "c")

```


<br>

# Demographic Data Frequencies:

```{r demographic data, echo=TRUE}

# Create summary table with counts and percentages with gtsummary
patient %>%
  select( #Select variables
    race, ethnicity, financialclass,
    hypertension, CHF, diabetes
  ) %>%
  tbl_summary(label = list( #Clarify variable names for output
    race ~ "Race",
    ethnicity ~ "Ethnicity",
    financialclass ~ "Financial Class",
    hypertension ~ "Hypertension",
    CHF ~ "Congestive Heart Failure",
    diabetes ~ "Diabetes"
  ))
```
