---
title: "721 Final"
author: "Caroline Falvey"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: show
---

```{r setup, include=TRUE, echo = TRUE, message = FALSE, results = "hide"}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(knitr)
library(here)
library(Hmisc)
library(ggplot2)
library(tidyverse)
library(flextable)
library(patchwork)
library(here)

# Set WD
here::here("Duke", "Fall 2022", "721", "Final Project")
# Read in simulation, table, and plotting functions
source("finalfunctions_Falvey.R")
```
<br>

## Part I:
<br>

```{r, echo = TRUE}
# Initialize sample size, thetas, & # of simulations and set seed
samp_size <- c(20, 40, 100)
theta <- c(0.05, 0.15)
runs <- 10000
set.seed(123456789)

# Create vector for dimension names in the NA arrays
dimnames <- list(samp_size, c("exact", "wilson", "asymptotic"), theta)

# Initialize array for coverage probability
coverage_array <- array(NA,
  dim = c(
    length(samp_size), 3,
    length(theta)
  ),
  dimnames = dimnames
)

# Initialize array for interval width
width_array <- array(NA,
  dim = c(
    length(samp_size), 3,
    length(theta)
  ),
  dimnames = dimnames
)


# Loop through sample sizes & theta options to fill in arrays
# Rows = sample sizes, columns = interval type, arrays = theta values
for (i in seq_along(samp_size)) {
  for (j in seq_along(theta)) {
    data <- simulation(samp_size[i], theta[j], runs)

    coverage_array[i, , j] <- data$coverage_prob
    width_array[i, , j] <- data$width
  }
}


```

<br>

```{r Tables, echo = TRUE}
# Create flextables to output results:

# Table for coverage probability
coverage_table <- table_fnt(
  array = coverage_array,
  theta = theta,
  samp_size = samp_size,
  caption = "Coverage Probability"
)

# Table for interval width
width_table <- table_fnt(
  array = width_array,
  theta = theta,
  samp_size = samp_size,
  caption = "Interval Width"
)

```
<br>

`r {coverage_table}`

<br>

`r {width_table}`

<br>

```{r Graphs, echo = TRUE, results = TRUE}
# Create plots

# Plot for coverage probabilities
coverage_plot <- plotter(
  array = coverage_array,
  samp_size = samp_size,
  theta,
  caption = "Coverage Probability"
)

# Plot for interval widths
width_plot <- plotter(
  array = width_array,
  samp_size = samp_size,
  theta = theta,
  caption = "Interval Width"
)

# Output plots using patchwork library
coverage_plot / width_plot

```
<br>



## Part II:

Based on the simulation results from Part I above, we will use the exact interval. The exact interval consistently has the highest coverage probability across the three sample sizes in our simulation. Additionally, although the exact test consistently has the widest interval, the width is not significantly larger when compared to the other tests and it converges to 0 as the sample size gets larger. 

<br>

## Part III:
```{r, echo = TRUE}
# Read in mortality data
mortality_data <- read_csv("mortality_data.csv", na = "-999")

# Data Cleaning
mortality_clean <- mortality_data |>
  na.omit() |> # Remove NAs
  filter(sex == "M" | sex == "F") |> # make sure only sexes are M/F
  filter(death == "0" | death == "1") # make sure only death outcome is binary

# Table of mortality by sex
mortality_summary <- addmargins(table(mortality_clean$sex,
                                      mortality_clean$death))

# Initialize NA matrix to store point estimates & intervals
all_summary <- matrix(NA, nrow = nrow(mortality_summary), ncol = 3,
                      dimnames = list(c("Female", "Male", "Total"), 
                                      c("pt_est", "low", "upp")))

# Calculate pt estimates & CI for each sex by looping through rows of the mortality table
for (i in 1:3){
  all_summary[i, ] <- binconf(mortality_summary[i, 2],
                              mortality_summary[i, 3],
                              method = "exact")
}

# Create a table to output displaying estimates and CIs
summary_table <- kable(all_summary,
                       col.names = c("Estimate", "Lower Bound", "Upper Bound"),
                       align = "c",
                       caption = "95% Exact Confidence Interval for Mortality Rate",
                       digits = 3)
```
<br>
`r {summary_table}`